Capturing the shape of our experiences and how we remember them

### 2018-08-20 paper/main.tex
checking in edits

Human memory studies typically assess memory for the elements within an experience, such as the people, places and things encountered. However, real life experiences can also be described in terms of their temporal structure (i.e. how the contents relate/change over time).  Naturalistic experiences are (typically) autocorrelated in space and time on short timescales: the contents of an experience are often similar from moment to moment and change gradually. For example, many of us spend our mornings/evenings at home and much of our days at work (e.g. an office). While the specific happenings within these contexts differ, they are often highly correlated (and thus predictable) over time. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The spatial contexts before and after lunch are highly similar, but separated in time. Thus, while evaluating the ability to recall specific contents of an experience has been a fruitful way to learn about human memory, another important dimension to consider is the temporal structure of the remembered experience, and how the temporal structure of the memory relates to the original experience. However, these recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely critical if we wish to understand how our memory systems remember our life experiences.

While our life experiences often change gradually, they can also change rapidly (e.g. walking through a doorway) \citep{Radv12}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize/organize our memories into ``events'' \citep{Radv12, RadvZack11, BrunEtal18, HeusEtal18, ClewDava18}. Short timescale autocorrelations in the contents of an experience are thought associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, Radv12}. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people naturally recount their experiences in a way that is structured around events is not known.

In this study, we analyzed an open dataset in which participants view and then verbally recall an episode of the BBC series \textit{Sherlock} \citep{ChenEtal17}. To capture the flow of the video's content (i.e. the storyline), we developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17}. This model casts naturalistic stimuli as high-dimensional topic ``trajectories'', whereby the theme of each moment of the video is represented as a topic vector. Successful remembering entails verbally ``traversing'' the path formed by the stimulus topic trajectory as closely as possible. We hypothesized that successful memory for a naturalistic stimulus is given by a recall trajectory that 1) has temporal structure similar to the stimulus 2) closely follows the path of the stimulus trajectory and 3) covers the majority of the stimulus path.

@@

\subsubsection{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters, location, and a short summary of the scene (see Fig.\ref{fig:schematic} for example text, see \ref{sec:methods} for analysis details). As depicted in Fig.~\ref{fig:model}a, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic `state' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}c).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly descriminable.

After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. An example participant's (\#13) recall model is plotted in Fig.~\ref{fig:model}b. Like the video model, topic vectors were sparse and changed gradually.  Then, we investigated the temporal structure of the recall matrices. From each participant's recall model, we computed a timepoint (68 to 294) by timepoints by timepoints (range: 68 to 294) correlation matrix. An example participant's (\#13) correlation matrix is plotted in Fig.~\ref{fig:model}d). Like the video model, each participants' recall correlation matrix exhibited a strong block-diagonal structure (Supp. Fig.~\ref{fig:corrmats})). Notably, this suggests that participants recounted the video in discriminable segments, likely related to the recall of specific events from the video.

@@

\subsubsection{Segmenting the video and recall models into `events'}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}c,d).  We hypothesized that this structure might arise from transient stability in "theme" of the video/recall (i.e. ``events''). We reasoned that specific events described by a participant could be ``matched'' to specific video events by computing the similarity between their topic vectors. In other words, the video event that is most similar (i.e. correlated) to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e. states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the model and a representative participant (\#13) are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}b,d).

Next, we created a video ``event model'' by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}e).  We performed the same procedure for the recall matrices (see Fig.~\ref{fig:model}f for example). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig~\ref{fig:model}g). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we computed the index of the video event with the highest correlation to the recall event (i.e. the argmax).  This is depicted in Fig.~\ref{fig:model}g as the cells highlighted with a yellow border. Notably, our algorithm suggests that the example participant recalled most of the video events and did so in order.

@@

\subsubsection{The shape of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization. However, none of these approaches capture the rich autocorrelated structure inherent in naturalistic stimuli and associated recall of those experiences. In the next set of analyses, we test the hypothesis that successful recall of a naturalistic stimulus entails recovering the trajectory of a stimulus.

@@

To visualize the relationship between the video and recall event models, we embedded them into a 2D space (using the UMAP dimensionality reduction algorithm, for details see \ref{sec:methods}) where the points represent video/recall events and the distance between them represents their similarity/distance in ``topic space'' (Fig.~\ref{fig:trajectory}a).  We observed that visually, it appears that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space''. For each point on the grid, we drew a circle (radius=.25) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test (p$<$.001, corrected at p$<$.05 using permutation procedure described in Methods). Thus participants' recalls followed the same path as the video model (Fig.~\ref{fig:trajectory}b), suggesting the shape of the stimulus was preserved despite idiosyncratic differences between participants in their recalls.

\subsubsection{Additional measures of naturalistic recall}
Representing the video and verbal recall as events in ``topic space'' also affords us the ability to characterize the quality of recall in a more fine-grained and nuanced way than what was previously possible. In addition to the shape analyses described above, we quantified the distance between each video event and its matching recall event (``precision''), the the distance between a video event and all non-matching recall events (``distinctiveness'') and the temporal order of the recall events (``order''). All three of these metrics were correlated to hand-annotated memory performance across participants ($p$<.05, see Supplemental Materials Section 1 for details).

Just like in a traditional ``free-recall'' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. We considered 6 classic list-learning analyses: overall accuracy, probability of first recall, lag-conditional response probability, serial position curve, temporal clustering and semantic clustering. Extending these classic list-learning analyses to naturalistic stimuli allows to directly compare trial- and naturalistic-based recall studies (see Supplemental Materials Section 2 for details).

@@

\subsubsection{Memorability by event and topics}
A notable advantage of representing naturalistic stimuli/memory using this approach is that the event models can be mapped back to the language used to fit the model(s).  This allows for analysis and visualization of the language used when participant's recounted their experiences.  In this next analysis, we 1) quantified the memorability of each video event and 2) used memorability to plot the top words for the most and least memorable scenes, and also aggregated across the entire video.

To measure video event memorability, we computed the correlation between each video event and the closest recall event and then averaged those values within participant (Fig.~\ref{fig:topics}a). We used event memorability to create a weighted average over all video event vectors, where video events with better memory were weighted more heavily. Then, we extracted the top 200 words and created ``wordles'' representing the most memorable themes in the video. The result reveal that scenes containing ``Sherlock'' and ``John'' were highly memorable, and also scenes containing words such as ``medium'', ``street'' and ``baker'' (Fig.~\ref{fig:topics}b). To find words related to the least memorable scenes, we inverted the memorability weights and again extracted the top 200 words.  Scenes containing ``Mike'' and ``Molly'' were least remembered, and also words like ``conference'', ``hospital'' and ``psychotherapist'' (Fig.~\ref{fig:topics}b).

@@

@@

Historically, human memory reseachers have focussed on measuring the quantity of information recovered by an individual (i.e. the proportion correct) under different experimental conditions. However, for real world experiences this is not a well-defined task. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience. Our work characterizes experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the movie participants viewed by measuring the match between the shapes of the movie's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory-- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape.

@@

\textbf{Temporal clustering.} Temporal clustering measures the extent to which participants group their recall responses according to encoding position. For instance, if the participant recalled each item in the presentation order, this would result in a score of 1. If the participant recalled randomly with respect to presentation order, this would result in a score of ~.5.  For each event transition (and separately for each participant), we computed the rank similarity (euclidean distance) between the presentation position  of the current and next recall events. The scores were then averaged within participant to get a single number representing the extent of temporal clustering exhibited by a given participant.

\textbf{Semantic clustering.} Similar to temporal clustering, semantic clustering measures the extent to which participants group their recall responses according to semantic similarity. Here, we are using the topic vectors for each event as a proxy for its semantic content. Thus, similarity between the semantic content for two events can be computed by correlating their respective topic vectors.  For instance, if each consecutive recall was the next most similar event (in terms of its s), this would result in a score of ~1. If the participant recalled randomly with respect to semantic similarity, this would result in a score of ~.5.  For each event transition (and separately for each participant), we computed the rank similarity (correlation distance) between the current recall event and the next recall event. The scores were then averaged within participant to get a single number representing the extent of semantic clustering exhibited by a given participant.

>>>>

Human memory studies typically assess memory for the elements within an experience, such as the people, places and things encountered. However, real life experiences can also be described in terms of their temporal structure (i.e. how the contents relate/change over time).  Naturalistic experiences are autocorrelated in space and time on short timescales: the contents of an experience are often similar from moment to moment and change gradually. For example, many of us spend our mornings/evenings at home and much of our days at work (e.g. an office). While the specific happenings within these contexts differ, they are often highly correlated (and thus predictable) over time. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The spatial contexts before and after lunch are highly similar, but separated in time. Thus, while evaluating the ability to recall specific contents of an experience has been a fruitful way to learn about human memory, another important dimension to consider is the temporal structure of the remembered experience, and how the temporal structure of the memory relates to the original experience. However, these recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely critical if we wish to understand how our memory systems remember our life experiences.

While our life experiences often change gradually, they can also change rapidly (e.g. walking through a doorway) \citep{Radv12}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{Radv12, RadvZack11, BrunEtal18, HeusEtal18, ClewDava18}. Short timescale autocorrelations in the contents of an experience are thought associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, Radv12}. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people naturally recount their experiences in a way that is structured around events is not known.

In this study, we analyzed an open dataset in which participants view and then verbally recall an episode of the BBC series \textit{Sherlock} \citep{ChenEtal17}. To capture the temporal structure of the video, we developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17}. This model casts naturalistic stimuli as high-dimensional topic ``trajectories'', whereby the theme of each moment of the video is represented as a topic vector. Successful remembering entails verbally ``traversing'' the path formed by the stimulus topic trajectory as closely as possible. We hypothesized that successful memory for a naturalistic stimulus is given by a recall trajectory that closely follows the path (and covers the entirety) of the stimulus trajectory.

@@

\subsubsection{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters, location, and a short summary of the scene (see Fig.~\ref{fig:schematic} for example text, see Methods for analysis details). As depicted in Fig.~\ref{fig:model}a, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic ``state'' to the next (possibly indexing scene transitions in the stimulus). To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}c).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly descriminable.

After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. An example participant's (\#13) recall model is plotted in Fig.~\ref{fig:model}b. Like the video model, topic vectors were sparse and changed gradually.  Then, we investigated the temporal structure of the recall matrices. From each participant's recall model, we computed a timepoint (68 to 294) by timepoints by timepoints (range: 68 to 294) correlation matrix. An example participant's (\#13) correlation matrix is plotted in Fig.~\ref{fig:model}d. Like the video model, each participant's recall correlation matrix exhibited a strong block-diagonal structure (Supp. Fig.~\ref{fig:corrmats})). Notably, this suggests that participants recounted the video in discriminable segments, likely related to the recall of specific events from the video.

@@

\subsubsection{Segmenting the video and recall models into `events'}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}c,d).  We hypothesized that this structure might arise from transient stability in the "theme" of the video/recall. We reasoned that specific events described by a participant could be ``matched'' to specific video events by computing the similarity between their topic vectors. In other words, the video event that is most similar (i.e. correlated) to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e. states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the video model and a representative participant's recall model are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}b,d).

Next, we created a video ``event model'' by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}e).  We performed the same procedure for the recall matrices (Fig.~\ref{fig:model}f). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig~\ref{fig:model}g). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we computed the index of the video event with the highest correlation to the recall event (i.e. the argmax).  This is depicted in Fig.~\ref{fig:model}g as the cells highlighted with a yellow border. Notably, our algorithm suggests that the example participant recalled most of the video events and did so in order.

@@

\subsubsection{The shape of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled \citep{Murd62a}. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization \citep{Kaha96, }. However, none of these approaches capture the rich autocorrelated structure inherent in naturalistic stimuli and associated recall of those experiences. In the next set of analyses, we test the hypothesis that successful recall of a naturalistic stimulus entails recovering the trajectory of a stimulus.

@@

To visualize the relationship between the video and recall event models, we embedded them into a 2D space (using the UMAP dimensionality reduction algorithm, for details see Methods) where the points represent video/recall events and the distance between them represents their similarity/distance in ``topic space'' (Fig.~\ref{fig:trajectory}a).  We observed that visually, it appears that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space''. For each point on the grid, we drew a circle (radius=.25) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test (p$<$.001, corrected at p$<$.05 using permutation procedure described in Methods). Thus participants' recalls followed the same path as the video model (Fig.~\ref{fig:trajectory}b), suggesting the shape of the stimulus was preserved despite idiosyncratic differences between participants in their recalls.

@@

\subsubsection{Memorability by event and topics}
A notable advantage of representing naturalistic stimuli/memory using this approach is that the event models can be mapped back to the language used to fit the model(s).  This allows for analysis (and visualization) of the language used when participants' recounted their experiences.  In this next analysis, we 1) quantified the memorability of each video event and 2) used memorability to plot the top words for the most and least memorable scenes.

To measure video event memorability, we computed the correlation between each video event and the closest recall event and then averaged those values within participant (Fig.~\ref{fig:topics}a). We used event memorability to create a weighted average over all video event vectors, where video events with better memory were weighted more heavily. Then, we extracted the top 200 words and created ``wordles'' representing the most memorable themes in the video. The results reveal that scenes containing ``Sherlock'' and ``John'' were highly memorable, and also scenes containing words such as ``medium'', ``street'' and ``baker'' (Fig.~\ref{fig:topics}b). To find words related to the least memorable scenes, we inverted the memorability weights and again extracted the top 200 words.  Scenes containing ``Mike'' and ``Molly'' were least remembered, and also words like ``conference'', ``hospital'' and ``psychotherapist'' (Fig.~\ref{fig:topics}b).

@@

\subsubsection{Additional measures of naturalistic recall}
Representing the video and verbal recall as events in ``topic space'' also affords us the ability to characterize the quality of recall in a more fine-grained and nuanced way than what was previously possible. We quantified the distance between each video event and its matching recall event (``precision''), the the distance between a video event and all non-matching recall events (``distinctiveness'') and the temporal order of the recall events (``order''). All three of these metrics were correlated to hand-annotated memory performance across participants (p$<$.05, see Supplemental Materials Section 1 for details).

Just like in a traditional ``free-recall'' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. We considered 6 classic list-learning analyses: overall accuracy, probability of first recall \citep{Hoga75, HowaKaha99; Lami99}, lag-conditional response probability \citep{Kaha96, HowaKaha99}, serial position curve \citep{Murd62a}, temporal clustering and semantic clustering \citep{HowaKaha02, PolyEtal09}. Extending these classic list-learning analyses to naturalistic stimuli allows to directly compare trial- and naturalistic-based recall studies (see Supplemental Materials Section 2 for details).

@@

Here, we find that participants Historically, human memory reseachers have focussed on measuring the quantity of information recovered by an individual (e.g. the ``proportion correct'') under different experimental conditions. However, for real world experiences this is not a well-defined task. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience. Our work characterizes experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the movie participants viewed by measuring the match between the shapes of the movie's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory-- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape.

@@

\textbf{Temporal clustering.} Temporal clustering measures the extent to which participants group their recall responses according to encoding position ~\citep{PolyEtal09}. For instance, if the participant recalled each item in the presentation order, this would result in a score of 1. If the participant recalled randomly with respect to presentation order, this would result in a score of ~.5.  For each event transition (and separately for each participant), we computed the rank similarity (euclidean distance) between the presentation position  of the current and next recall events. The scores were then averaged within participant to get a single number representing the extent of temporal clustering exhibited by a given participant.

\textbf{Semantic clustering.} Similar to temporal clustering, semantic clustering measures the extent to which participants group their recall responses according to semantic similarity ~\citep{PolyEtal09}. Here, we are using the topic vectors for each event as a proxy for its semantic content. Thus, similarity between the semantic content for two events can be computed by correlating their respective topic vectors.  For instance, if each consecutive recall was the next most similar event (in terms of its s), this would result in a score of ~1. If the participant recalled randomly with respect to semantic similarity, this would result in a score of ~.5.  For each event transition (and separately for each participant), we computed the rank similarity (correlation distance) between the current recall event and the next recall event. The scores were then averaged within participant to get a single number representing the extent of semantic clustering exhibited by a given participant.
